Human capital mediates natural selection in contemporary humans

### 2021-12-01 why-natural-selection.Rmd
Focus on motivating the theory as a reason for sample-splitting. Reorder intro to put theory first.

As yet we know little about the social mechanisms behind these effects. This
study uses data from UK Biobank [@bycroft2018uk] to learn more. We test for
natural selection on 33 different polygenic scores by estimating their
correlation with fertility. We extend the analysis over two generations, using
data on respondents' number of siblings as well as their number of children.
This is interesting because consistent natural selection over multiple
generations could lead to substantive effects in the long run. Most importantly,
we examine reproductive rates in different subgroups of the population, in order
to uncover patterns that can help illuminate the mechanisms behind modern
natural selection.

We find selection effects on many polygenic scores. Effects are largely consistent
across generations. The strength of natural selection on a polygenic score is
associated with that score's correlation with education and earnings: scores
that predict lower education and earnings are being selected for. Also, across
the board, polygenic scores have stronger relationships with fertility among
specific subgroups. Selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with more lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed: polygenic scores predicting
higher education and earnings are associated with *higher* fertility.

These patterns are in line with economic theories of fertility
[@becker1960economic]. In these, higher potential earnings have two opposite
effects on fertility: a fertility-increasing *income effect* (higher income
makes children more affordable), and a fertility-lowering *substitution effect*
(time spent on childrearing has a higher cost in foregone earnings). Our results
suggest that the substitution effect dominates for single parents and younger
parents, while among couples and older parents (22+) effects are more evenly
balanced. Thus, contemporary natural selection on polygenic scores can be
explained by their correlation with earnings-increasing human capital. Below, we 
show that a simple model of human capital, education and fertility choices 
can give rise to our key empirical results.

@@

To understand the reasons for the ascertainment bias in our sample, and to learn 
about the mechanisms underlying natural selection, we split the sample, starting with
basic demographic variables including education, income and sex. These are all
potential sources of ascertainment bias, as mentioned above. We then re-estimate
\@ref(eq:regression) within each subgroup.

We emphasize that this exercise does
*not* apportion the total correlation with fertility into subgroups. 
That is because the polygenic score may also affect the probability of being in 
each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
fertility among university graduates, and among non-graduates; it also changes
a person's chance of attending university. Nevertheless, differences in equation
\@ref(eq:regression) across subgroups can inform us about how natural
selection works, by constraining the set of possible mechanisms.

@@

We rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
and using our age/qualification weights. Scores for hip circumference and drinks
per week show significant balancing selection ($p < 0.05/33$, negative
coefficient). Scores for educational attainment (EA2 and EA3) show significant
diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
strength of selection against educational attainment at very high levels of the 
PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

>>>>

As yet we know little about the social mechanisms behind these effects. A
potential explanation for them comes from the economic theory of fertility first
presented by [@becker1960economic]. In this theory, higher potential earnings
have two opposite effects on fertility: a fertility-increasing *income effect*
(higher income makes children more affordable), and a fertility-lowering
*substitution effect* (time spent on childrearing has a higher cost in foregone
earnings). Thus, an individual's *human capital* -- skills and personality
traits which are valuable in labour markets -- can increase or decrease his or
her fertility. This can lead to selection on inherited
variants which are linked to human capital. Furthermore, the economic theory predicts
systematic variation in the relative strength of the income and substitution
effects across different social groups.

This
study uses data from UK Biobank [@bycroft2018uk] to learn more about contemporary
natural selection. We test for
natural selection on 33 different polygenic scores by estimating their
correlation with fertility. We extend the analysis over two generations, using
data on respondents' number of siblings as well as their number of children.
This is interesting because consistent natural selection over multiple
generations could lead to substantive effects in the long run. We then examine
reproductive rates in different subgroups of the population. 
We find selection effects on many polygenic scores. Effects are largely consistent
across generations.

Next, we examine correlations with fertility in different subgroups. Across
the board, selection effects are stronger in groups with lower
income and less education, among younger parents, people not living with a partner,
and people with more lifetime sexual partners. Outside these groups, effects are
weaker and often statistically insignificant. In some subgroups, the
direction of selection is even reversed.

We then show that a parsimonious model of human capital, education and fertility
choices can give rise to these empirical results. To test the theory, we show
that polygenic scores' correlation with fertility is associated with their
correlation with education and earnings. Furthermore, controlling for education
and earnings reduces the correlation with fertility, suggesting that human
capital mediates natural selection. Thus, contemporary natural selection on
polygenic scores can be explained by their correlation with earnings-increasing
human capital.

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related 
polygenic scores. Nevertheless, differences in equation
\@ref(eq:regression) across subgroups constrain the set of possible explanations
for natural selection. As we show below, the economic theory predicts 
the patterns in the data, while alternative explanations do not.

@@

Balancing selection reduces variance in the trait under selection, 
while diversifying selection increases variance. To check for these,
we rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
and using our age/qualification weights. Scores for hip circumference and drinks
per week show significant balancing selection ($p < 0.05/33$, negative
coefficient). Scores for educational attainment (EA2 and EA3) show significant
diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
strength of selection against educational attainment at very high levels of the 
PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.
